Direct observation of Feshbach enhanced \swave scattering of fermions

### 2015-07-09 draft1.tex
Another round of Ian's comments - in progress

\begin{document}
\newcommand{\swave}[0]{$\it{s}$-wave }
\newcommand{\pwave}[0]{$\it{p}$-wave }
\newcommand{\K}{ $^{40}\rm{K}$}
\newcommand{\Rb}{ $^{87}\rm{Rb}$}
\newcommand{\us}{$\mu \rm{s}$} 
\newcommand{\mT}{$m\rm{T}$}
\newcommand{\ez}{$\hat{e}_z$}
\newcommand{\um}{$\mu \rm{m}$}

@@

\begin{abstract}
We directly measured the relative \swave{} scattering cross-section of ultracold \K{} atoms accross the 20.2 \mT{} Feshbach resonance. We collided a pair of degenerate Fermi clouds and imaged the scattered atoms. Owing to their low density, few atoms scattered, even near the resonance. To optimize signal to noise, we developed techniques to interpret absorption imaging in the regime where the optical intensity changes dramatically as light traverses the cloud, and recoil induced 
detuning corrections are significant. By applying these techniques to our \swave{} scattering data we were able to extract the resonant magnetic
field value  $B_0 = 20.247(2)$  \mT{} and resonance width $\Delta = 1.0(1)$  \mT{}. These imaging techniques are generally applicable to experimets requiring accurate detection of low atomic densities.

\end{abstract}

@@

Feshbach resonances are a commonly used technique for tuning the interaction strength in ultracold atomic gases. They have been particularly instrumental in the study of interactions and interaction-dependent processes in cold Fermi gases. Even weak interactions play a crucial role in the physics of atomic Bose-Einstein Condensates (BECs), for example giving rise to their characteristic Thomas-Fermi density profiles \cite{KetterleBEC}. The effects of interactions in Fermi gases, however, are harder to observe. The density of Fermi clouds is typically $\sim1000$ times less than that of BECs, making it necessary to enhance the strength of interactions in order to observe them \cite{KetterleDFG}. The tunability of interactions provided by Feshbach resonances has allowed for creation of molecular Bose-Einstein condensates from Fermi gases \cite{Greiner03,Zwierlein03, Jochim03} as well as observation of the phase transition from the BCS to BEC regime at sufficiently low temperatures \cite{Bartenstein04, Bourdel04, Zwierlein04, Regal04}. 
\par A Feshbach resonance occurs when a few-atom bound molecular state energetically approaches the free atomic state \cite{Chin10, Timmermans99}. The relative energy of the free atomic states and the molecular state is defined by a bias magnetic field. The Feshbach resonance can thus be approached by changing the bias field. In the simple case where there are no inelastic two-body channels, which is accurate for \K{} resonance discussed in this work, the effect of the resonance on the scattering length between two free atoms is \cite{Chin10}

@@

where $a$ is the scattering length, $a_{bg}$ is the scattering length away from the resonance, $\Delta$ is the width of the resonance, and $B_0$ is the field value at which the resonance occurs. Note that the scattering length tends to infinity from either side of the resonance.
\par  The exact value of the resonant field $B_0$ is difficult to calculate analytically and is commonly estimated via numerical models \cite{Tiesinga93, Lysebo09, Gao11} or determined experimentally \cite{Inouye98, Cornish00}. Many experimental techniques have been used to characterize Feshbach resonances, including the observation of atom loss due to three-body inelastic scattering, measurement of re-thermalization timescales, and imagin anisotropic expansion of the cloud upon release from a confining potential, all of which infer the elastic scattering cross section from collective behavior of the cloud \cite{Regal03,OHara02,Monroe93}. 
\par Here we present a new technique for measuring the location and width of a Feshbach resonance. We collided a pair of ultra-cold Fermi gas clouds and directly imaged the resulting \swave scattering halo as a function magnetic field strength. This allowed us to observe the enhancement in scattering without relying on proxy effects. We measured the fraction of atoms scattered during the collision, and from this fraction deduced the resonant magnetic field  and width of the resonance. A similar technique has been used to characterize impurity scattering in BECs \cite{Chikkatur00}.

@@

\par In our dilute Fermi clouds, even with the resonant enhancement of the scattering cross section, only a small fraction of the atoms underwent a scattering event as the clouds passed through each other. This made direct detection of \swave scattering halos difficult due to detection uncertainty, which disproportionately affected regions of low atomic density in absorption images. To optimize the signal to noise for low atom numbers, we utilized a non standard imaging regime. In this regime, the atoms acquire a non-negligible velocity during imaging and are Doppler-shifted away from the light field used to detect them.  Simulation of the absorption imaging process was necessary for an accurate interpretation of these images. With the simulation corrected images, we were able to extract the number and the scattered fraction of the atoms.
\par This paper is in two parts. In the first, we study absorption imaging in the presence of significant time-dependent Doppler shift and show how we use our results to interpret data. In the second, we describe our\swave scattering experiment and extract a measure of the location and width of the Feshbach resonance in \K{}.

@@

Absorption imaging is commonly thought of as the atomic cloud casting a "shadow" on a camera when illuminated by laser light. This imaging scheme relies on optical transitions between the ground and certain excited atomic states. Such atomic transitions have an energy difference $E_0$ with an associated frequency $\omega_0 = E_0/\hbar$, and a natural transition linewidth $\Gamma$. When interacting with a laser light field an atom will scatter photons: it will absorb a photon from the light field and move up to its excited state,  then re-emit the photon and decay back to its ground state. The  rate at which this scattering occurs, in the two-level atom approximation, is \cite{LCT}

@@

where $I/I_{\rm{sat}}$ is the laser intensity as a fraction of the saturation intensity, and $\delta$ is the detuning, the difference between the resonant transition frequency $\omega_0$ and the frequency of the laser light $\omega_L$.  
\par An absorption image is obtained by shining an on or near resonant probe beam ($\delta\approx0$) onto the atomic cloud. Some of the light is scattered by the atoms, and the unscattered light, $I_f(x,y)$, is imaged onto a camera, as seen in Fig. \ref{fig:absorptionIntor}a (top). After the atoms leave the imaging volume, the probe light is reapplied to calibrate the intensity $I_0(x,y)$ of light unaffected by the atoms (bottom). 

@@

\par The intensity $I_f(x,y)$ is related to the number of atoms the light encountered. Consider the light as it travels along the imaging axis $e_z$ through a 3-D atomic density profile $\rho(x,y,z)$. For the purposes of this discussion, let us focus on a single pixel of the camera, and thus a single value of $I_0$ and $I_f$, and a single column of atoms, $\rho(z)$. We are not able to infer the entire atomic distribution, but for each pixel we can obtain the integrated density, $n = \int \rho\left(z\right) \,\mathrm{d}z$. As the light travels through a column of atoms, each atom scatters light according to Eq. (\ref{eq:scatrate}). Therefore, the atoms further along the imaging axis $z$ experience a reduced optical intensity due to attenuation by the other atoms, as seen in Fig. \ref{fig:absorptionIntor}b. On resonance ($\delta=0$), the intensity change from scattering as a function of $z$ is

@@

\frac{dI(z)}{dz}=\hbar\omega_L\rho(z)\gamma_{sc}(z)=-\rho(z)\sigma_0\frac{I(z)}{1+I(z)/I_{\rm{sat}}},

@@

where $\sigma_0$ is the resonant scattering cross section. 
 This equation is integrated to obtain  \cite{Reinaudi07}

@@

\sigma_0 n = -\ln\left(\frac{I_f}{I_0}\right) + \frac{I_0-I_f}{I_{\rm{sat}}}.

@@

In the limit where the probe intensity is much smaller than the saturation intensity, $I_0\ll I_{\rm{sat}}$, this reduces to the optical depth, $OD_0=-\ln \left(I_f/I_0\right)$   \cite{Reinaudi07}. This is the simplest possible model that relates the observed intensities to the column density $n$. When the probe intensity is comparable to the saturation intensity, the second term in Eq. (\ref{eq2}) becomes significant, and we define the optical density corrected for high probe intensity $OD_1$ as the right hand side of Eq.  (\ref{eq2}).
\par There are further corrections that this equation does not take into account. In particular, it neglects the atomic recoil momentum and its effect on the laser detuning \cite{Konstantinidis12}. When an atom absorbs a photon from the laser light field and is excited to a higher energy level, by conservation of momentum it must also acquire a velocity increment in the direction of the light field. This recoil velocity is given by $v_r=\hbar k/m$, where $k=2\pi/\lambda$ is the wavenumber of the light field with wavelength $\lambda$ and $m$ is the atomic mass. When the atom returns to its ground state, the photon is re-emitted with some momentum $\vec{p}_e$. Over many photons, this momentum distribution averages to zero, $\int\vec{p}_e\mathrm{d}^3 x=0$. Therefore, on average, the atom  will acquire a net velocity of $v_r$ in the \ez{} direction.  The variance of $\vec{p}_e$, however, is not zero, allowing the atoms to acquire some momentum transverse to the laser field. We will ignore this correction, but the effect of this on the atomic cloud is pictured in Fig. \ref{fig:absorptionIntor}c. 
\par Once the atoms absorb enough photons that they acquire a substantial velocity in the direction of the light field, this velocity Doppler shifts them away from resonance with the light field. After scattering $N$ photons an atom acquires an average velocity $N v_r$ and an additional detuning $\delta=k N v_r$. Therefore, even if the probe beam is initially on resonance with the atomic transition, we cannot neglect the detuning term in the scattering rate as time goes on. Furthermore, this detuning varies both with imaging time $t$ and with distance along the propagation direction $z$ (Fig. \ref{fig:expos}a). Thus, the intensity lost to the atoms also acquires a time dependence: 

@@

\frac{dI(t,z)}{dz}=\sigma_0 \rho \frac{I(t,z)}{1+[2\delta(t,z)/\Gamma]^2 +I(t,z)/I_{\rm{sat}}}, \label{eq3}

@@

where the detuning $\delta$ is

@@

\delta(t,z)=\frac{v_r}{\hbar c \rho}\int_0^t \frac{dI(z,\tau)}{dz}\,\mathrm{d}\tau; \label{eq4} 

@@

\caption{(a) Dependence of velocity and detuning on position along the imaging axis in an atomic cloud of \K{} for three different imaging times and a probe intentisy $I_0=1.5 I_{\rm{sat}}$, as obtained by numerical simulation. (b) Using time dependent $I_f$ values obtained from recoil detuning corrected simulation of on-resonant imaging of $^{40}K$ atoms at probe intensity $I_0=0.8\, I_{\rm{sat}}$, this graph shows the optical depths obtained by each model. The `true' optical depth is $\sigma_0 n=1.6$. $OD_1$ is the high probe intensity corrected optical depth given by Eq. (\ref{eq2}). $OD_2$ is the high probe intensity corrected and expanded to second order in time optical depth, Eq. (\ref{eq:OD2}) \cite{LJLthesis}. The recoil time is the time it takes for the cloud, on average, to become detuned by a linewidth $\Gamma$. Both models start to differ from the true value before a recoil time.}  

@@

\par By considering this equation perturbatively in time we obtain corrections to second order in imaging time \cite{LJLthesis}

@@

\sigma_0n\approx OD^{(0)}+OD^{(1)}t+OD^{(2)}t^2 = OD_2

@@

 However, as shown in  Fig. \ref{fig:expos}b, the perturbative treatment breaks down shortly after the zeroth order approximation of Eq. (\ref{eq2}). To adequately correct for the recoil induced detuning of the atoms, we numerically simulated the imaging process to obtain $I_f$ as a function of imaging time, atomic density, and probe intensity. 

@@

\par In the following, we describe two versions of this simulation. First, we took a simplistic approach where the spatial distribution of atoms does not change appreciably during the imaging time: $vt\ll I_0/(\hbar \omega_L \gamma_{sc} \rho)$. We tested this approach in known limits and then checked the validity of the static assumption. For realistic input parameters, we found this assumption to be invalid. We developed aquasi-classical approach and allowed the atoms to move during the imaging time, allowing us to simulate the phase space evolution of atoms subjected to probe light. While the atomic trajectories were wildly different than in the static atom approximation, the predicted $OD$s only varied on the 0.5$\%$ level between the two models.

>>>>

\begin{document}
\newcommand{\swave}[0]{$\it{s}$-wave }
\newcommand{\pwave}[0]{$\it{p}$-wave }
\newcommand{\K}{ $^{40}\rm{K}$}
\newcommand{\Rb}{ $^{87}\rm{Rb}$}
\newcommand{\us}{$\rm{\mu s}$} 
\newcommand{\mT}{$\rm{mT}$}
\newcommand{\ez}{$\bf{\hat{e}_z}$}
\newcommand{\um}{$\rm{\mu m}$}

@@

\begin{abstract}
We directly measured the \swave{} scattering cross-section of ultracold \K{} atoms accross the 20.2 \mT{} Feshbach resonance by colliding pairs of degenerate Fermi gas (DFG) clouds and imaging the scattered atoms. Owing to DFG's low density, few atoms scattered, even near the resonance. To optimize signal to noise, we developed techniques to interpret absorption images in the regime where recoil induced 
detuning corrections are significant. By applying these techniques to our \swave{} scattering data we extracted the resonant magnetic
field value  $B_0 = 20.247(2)$  \mT{} and resonance width $\Delta = 1.0(1)$  \mT{}. These imaging techniques are generally applicable to experimets requiring accurate interpretation of aborption images with a wide dynamic range of atomic densities. 
\end{abstract}

@@

Feshbach resonances are widely used for tuning the interaction strength in ultracold atomic gases. They have been particularly instrumental in the study of interactions and interaction-dependent processes in cold Fermi gases. In contrast to atomic Bose-Einstein condensates (BECs), where even weak interactions play a crucial role, for example giving rise to their characteristic Thomas-Fermi density profiles \cite{KetterleBEC}, the effects of interactions in Fermi gases are harder to observe. The density of Fermi clouds is typically $\sim1000$ times less than that of BECs, making it necessary to enhance the strength of interactions in order to observe them \cite{KetterleDFG}. The tunability of interactions provided by Feshbach resonances has allowed for creation of molecular Bose-Einstein condensates from Fermi gases \cite{Greiner03,Zwierlein03, Jochim03} as well as observation of the phase transition from the BCS to BEC regime at sufficiently low temperatures \cite{Bartenstein04, Bourdel04, Zwierlein04, Regal04}. 
\par A Feshbach resonance occurs when a two-atom bound molecular state energetically approaches the two-atom continuum \cite{Chin10, Timmermans99}. In experiment, the relative energy of the free atomic states and the molecular state is defined by a bias magnetic field, so the Feshbach resonance can be approached by changing the bias field. In the simple case where there are no inelastic two-body channels, such as for the \K{} resonance discussed in this work, the effect of the resonance on the scattering length between two free atoms is \cite{Chin10}

@@

 $a_{bg}$ is the bachground scattering length, $\Delta$ is the width of the resonance, and $B_0$ is the field value at which the resonance occurs.The scattering length diverges at the resonance.
\par  The exact value of the resonant field $B_0$ is difficult to calculate analytically and is commonly computed via numerical models based on experimental input parameters \cite{Tiesinga93, Lysebo09, Gao11} or determined experimentally \cite{Inouye98, Cornish00}. Many experimental techniques have been used to characterize Feshbach resonances, including the observation of atom loss due to three-body inelastic scattering, measurement of re-thermalization timescales, and anisotropic expansion of the cloud upon release from a confining potential, all of which infer the elastic scattering cross section from collective behavior of the cloud \cite{Regal03,OHara02,Monroe93}. 
\par Here we used direct scattering as a primary probe of the location and width of a Feshbach resonance. We collided pairs of ultra-cold Fermi gas clouds and directly imaged the resulting \swave scattering halos as a function magnetic field strength. This allowed us to observe the enhancement in scattering without relying on proxy effects. We measured the fraction of atoms scattered during the collision, and from this fraction deduced the resonant magnetic field  and width of the resonance. 
%A similar technique has been used to characterize impurity scattering in BECs \cite{Chikkatur00}.

@@

\par In our dilute Fermi clouds, even with the resonant enhancement of the scattering cross section, only a small fraction of the atoms scattered as the clouds passed through each other. This made direct detection of \swave scattering halos difficult due to detection uncertainty, that disproportionately affected regions of low atomic density in absorption images. To optimize the signal to noise for low atom numbers, we absorption imaged with fairly long, high-intensity pulses --- a non-standard regime, where the atoms acquired a velocity during imaging and the resulting Doppler-shift was non-negligible.  Simulation of the absorption imaging process was necessary for an accurate interpretation of these images. With the simulation corrected images, we extracted the fraction of atoms scattered in our collision experiment.
\par This paper is in two parts. In the first, we study absorption imaging in the presence of a significant time-dependent Doppler shift and show how we use our results to interpret data. In the second, we describe our\swave scattering experiment and extract a measure of the location and width of the Feshbach resonance in \K{}.

@@

Absorption imaginmeasures the shadow cast by an atomic ensemble in an illuminating probe laser beam on a camera. This imaging technique relies on optical transitions between the ground and certain excited atomic states. Such atomic transitions have an energy difference $\hbar\omega_0$, and a natural transition linewidth $\Gamma$. When interacting with a laser field an atom scatters photons from the laser field into the empty vacuum modes. The  rate at which this scattering occurs, in the two-level atom approximation, is \cite{LCT}

@@

where $I/I_{\rm{sat}}$ is the laser intensity as a fraction of the saturation intensity, and $\delta=\omega_L-\omega_0$ is the detuning, the difference between the resonant transition frequency and the frequency of the laser light.  
\par An absorption image is obtained by shining an on- or near-resonant probe beam ($\delta\approx0$) onto the atomic cloud. Some of the light is scattered by the atoms, and the unscattered light, $I_f(x,y)$, is imaged onto a camera, as depicted in Fig. \ref{fig:absorptionIntor}a (top). The probe light is reapplied with the atoms absent to calibrate the intensity $I_0(x,y)$ of light unaffected by the atoms (bottom). 

@@

\par The intensity $I_f(x,y)$ is related to the number of atoms the light encountered. Consider the light as it travels along the imaging axis \ez{} through a 3-D atomic density profile $\rho(x,y,z)$. For the purposes of this discussion, let us focus on a single pixel of the camera, and thus a single value of $I_0$ and $I_f$, and a single column of atoms, $\rho(z)$. We are not able to infer the entire atomic distribution, but for each pixel we may obtain the column density, $n = \int \rho\left(z\right) \,\mathrm{d}z$. As the light travels through a column of atoms, each atom scatters light according to Eq. (\ref{eq:scatrate}). Therefore, the atoms further along the imaging axis $z$ experience a reduced optical intensity due to attenuation of the laser field by the other atoms (Fig. \ref{fig:absorptionIntor}b). The intensity change from scattering as a function of $z$ is

@@

\frac{dI(z)}{dz}=\hbar\omega_L\rho(z)\gamma_{sc}(z)=-\rho(z)\sigma_0\frac{I(z)}{1+D^2 +I/I_{\rm{sat}}},
\label{eq:dIdz}

@@

where $\sigma_0$ is the resonant scattering cross section and we define a scaled detuning $D=2\delta/\Gamma$. Integrating this equation yields  \cite{Reinaudi07}

@@

\sigma_0 n = -\left(1+D^2\right)\ln\left(\frac{I_f}{I_0}\right) + \frac{I_0-I_f}{I_{\rm{sat}}}.

@@

When the probe intensity is much smaller than the saturation intensity, $I_0\ll I_{\rm{sat}}$, Eq. (\ref{eq2}) reduces to the optical depth, $OD_0=-\ln \left(I_f/I_0\right)$   \cite{Reinaudi07}. This is the simplest possible model relating the observed intensity to the column density $n$. When the probe intensity is comparable to the saturation intensity, the second term in Eq. (\ref{eq2}) becomes significant, and we define the optical density corrected for high probe intensity $OD_1$ as the right hand side of Eq.  (\ref{eq2}).
\par Equations (\ref{eq:dIdz})-(\ref{eq2}) neglect the atomic recoil momentum and its effect on the laser detuning \cite{Konstantinidis12}. When an atom absorbs a photon from the laser light field it acquires a momentum kick $\hbar k_R$ in the direction of the light field. The associated recoil velocity is $v_r=\hbar k_R/m$, where $k_R=2\pi/\lambda$ is the recoil momentum of the light field with wavelength $\lambda$ and $m$ is the atomic mass. The re-emitted photon imparts a similar recoil momentum $\vec{p}_e$. Over many scattering events this momentum distribution averages to zero. Therefore the atom  will acquire an average per-photon velocity $v_r$ along \ez{}.  The variance of $\vec{p}_e$, however, is not zero, allowing the atoms to acquire some momentum transverse to the laser field. We will ignore this correction, but the effect of this on the atomic cloud is pictured in Fig. \ref{fig:absorptionIntor}c. 
\par This velocity parallel the light field Doppler shifts the atoms from resonance. After scattering $N$ photons an atom acquires an average velocity $N v_r$ and an additional detuning $\delta=k N v_r$. Therefore, even if the probe beam is initially on resonance with the atomic transition, we cannot neglect the detuning's  effect on the scattering rate. Furthermore, this detuning varies both with imaging time $t$ and with distance along the propagation direction $z$ (Fig. \ref{fig:expos}a). Thus, the intensity lost to the atoms also acquires a time dependence: 

@@

\frac{dI(t,z)}{dz}=\sigma_0 \rho(t,z) \frac{I(t,z)}{1+D(t,z)^2 +I(t,z)/I_{\rm{sat}}}, \label{eq3}

@@

where the scaled detuning $D$ is

@@

D(t,z)=\frac{2 v_r}{\Gamma\hbar c \rho(t,z)}\int_0^t \frac{dI(z,\tau)}{dz}\,\mathrm{d}\tau; \label{eq4} 

@@

\caption{(a) Dependence of velocity and detuning on position simulated for \K{} at three different imaging times and a probe intensity $I_0=1.5 I_{\rm{sat}}$. (b) Optical depth from the time-averaged intensity $I_f$ obtained from recoil detuning corrected simulation of on-resonant imaging of $^{40}K$ atoms at probe intensity $I_0=0.8\, I_{\rm{sat}}$. The `true' optical depth is $\sigma_0 n=1.6$. $OD_1$ is the high probe intensity corrected optical depth given by Eq. (\ref{eq2}). $OD_2$ is the high probe intensity corrected and expanded to second order in time optical depth, Eq. (\ref{eq:OD2}) \cite{LJLthesis}.}  

@@

\par By considering this equation perturbatively in imaging time we obtain corrections to second order \cite{LJLthesis}

@@

\sigma_0n\approx OD^{(0)}+OD^{(1)}t+OD^{(2)}t^2\equiv OD_2

@@

As shown in  Fig. \ref{fig:expos}b, the perturbative treatment is accurate to times up to the recoil time, defined as the time it would take a single atom to be detuned by half an atomic linewidth, but diverges after that point. To adequately correct for the recoil induced detuning of the atoms, we numerically simulated the imaging process to obtain $I_f$ as a function of imaging time, atomic density, and probe intensity. 

@@

\par In the following, we describe two versions of this simulation. First, we took a simplistic approach where the spatial distribution of atoms did not change appreciably during the imaging time: $vt\ll I_0/(\hbar \omega_L \gamma_{sc} \rho)$. We verified this approach in known limits and then cross-checked the validity of the static assumption. For realistic input parameters, we found this assumption to be invalid. We then developed a quasi-classical approach and allowed the atoms to move during the imaging time. While the atomic trajectories were wildly different than in the static atom approximation, the predicted $OD$s only differed on the 0.5$\%$ level.
